% abstract.tex
The A=1 discrete causal lattice of Paper~I~\cite{menendez2026geometryfirst} singles
out an optical axis $\hat{\mathbf{n}}_* = (1,1,-1)/\sqrt3$---the one cube body-diagonal
absent from its three-vector bipartite hop set---and this single geometric fact governs
both the matter and the gauge sector along that axis. We report a clean split between
what survives and what is excluded as constructed.

\emph{What survives.} The lattice's plaquette geometry lands automatically on the
impedance-matched constitutive relation $\boldsymbol{\mu}^{-1} =
\gamma\operatorname{adj}(\boldsymbol{\varepsilon})$---an identity for \emph{any} hop
vectors---which forces the photon dispersion to a doubled transverse root. Given those geometric
plaquette blocks as the constitutive response, discrete three-hop vacuum electromagnetism
therefore \emph{cannot} be birefringent: the polarization split cancels identically
(numerically to $10^{-15}$), a hop-set-universal result immune to every deferred
normalization. (Establishing the blocks as the full matter-integrated dynamical action is
a separate, still-open step.) Read as a falsification test, the
observed absence of optical-axis vacuum birefringence is a stringent existing constraint
this class of models survives, while the CPT-odd, dimension-five and other birefringent
Lorentz-violating rivals are excluded by the same datum.

\emph{What is excluded as constructed.} The same single fixed hop set (one trigonal
domain) is a Lorentz violation in \emph{both} sectors. The induced photon carries an
order-unity, dimension-four common-mode speed anisotropy (factor $\sim2$ about
$\hat{\mathbf{n}}_*$), excluded by cavity/Michelson--Morley isotropy bounds. And the bare
single-tick spinor, read as literal laboratory matter, is worse: an \emph{exact} axial
flat band (axial transport identically zero), an order-unity directional kinetic
anisotropy, and an $\omega=0$ amplitude filter that accumulates \emph{linearly} in the
lattice spacing ($e$-folding times of days to minutes at Planckian $a$)---each excluded
by many orders. Both exclusions trace to the one arbitrary omission of the fourth
diagonal that fixes the axis, and both share their escape with the rest of the program:
an $O_h$-restoring (four-orientation) vacuum for the gauge loop---the same average by
which Papers~I--II obtain isotropic Maxwell and emergent Lorentz---and the
token$\to$matter emergence map for the spinor. Whether either construction preserves the
A=1 and Dirac structure is a foundational open question; we state the no-go plainly
rather than resolve it. Every claim's scope, evidence, and status is recorded in the
audit table (Table~\ref{tab:audit}).
