% ============================================================
% audit_table.tex
%
% A "system audit" table mapping the paper's claims to the evidence
% backing each one.  This is *the* canonical source of truth for which
% claims are PASS / PART / STUB / FAIL --- prose elsewhere in the paper
% should not contradict it.
%
% Status semantics (paste into the abstract or front-matter as well):
%   PASS  -- derivation complete or experiment confirms the claim to
%            stated precision.
%   PART  -- mechanism demonstrated, full quantitative match pending;
%            specific gaps documented in the evidence column.
%   STUB  -- analytical result with no numerical confirmation yet,
%            or planned experiment not yet run.
%   FAIL  -- tested and disconfirmed.
%
% Use \texttt{} for code/experiment names and file paths inside cells.
% Long cells are fine -- the column widths are tuned for prose.
% ============================================================

\label{tab:audit}

{\scriptsize
\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{0.95}
\begin{longtable}{@{}p{2.6cm}p{3.8cm}p{3.2cm}p{4.4cm}p{1.0cm}@{}}
\caption{\small System audit: paper claims mapped to supporting
derivations and experiments.  Status semantics in the front-matter
(\texttt{PASS} / \texttt{PART} / \texttt{STUB} / \texttt{FAIL}).}
\label{tab:audit_caption} \\
\hline
\textbf{Claim} & \textbf{Mechanism} & \textbf{Continuum / formal limit} & \textbf{Evidence} & \textbf{Status} \\
\hline
\endfirsthead

\multicolumn{5}{l}{\textit{(Table~\ref{tab:audit}, continued from previous page)}} \\
\hline
\textbf{Claim} & \textbf{Mechanism} & \textbf{Continuum / formal limit} & \textbf{Evidence} & \textbf{Status} \\
\hline
\endhead

\hline
\multicolumn{5}{r}{\textit{(continued on next page)}} \\
\endfoot

\hline
\endlastfoot

% ── Kinematic channel ───────────────────────────────────────────────
Kinematic birefringence along $(1,1,-1)$
    & Closed-form uniaxial split from the spinor propagator: $M_\text{eff} = \tfrac43(\mathbb{I}-\hat{\mathbf{n}}_*\hat{\mathbf{n}}_*^T)$, so $|H_\text{RGB}(\mathbf{k})|^2 = 1-\tfrac43|\mathbf{k}_\perp|^2$. Group-velocity split $\Delta v = \tfrac23\sin\omega\,|\mathbf{k}|$ (angular law $\sin\alpha$, massive sector) and $\omega$-independent dispersion-curvature split $4/3$ vs $0$ (photon + massive)
    & Isotropic $c$ under $O_h$ averaging / $a\to 0$
    & \emph{Analytic:} eigenphase dispersion $\tan\theta = \tan(\omega/2)/|H_\text{RGB}|$ and group velocity in closed form; $|H_\text{RGB}|^2 = 1$ exactly along $(1,1,-1)$ at all orders. \texttt{exp\_02\_closed\_form\_split\_check}: structure-factor and dispersion identities to $<10^{-12}$, speed coefficients to $<10^{-5}$. \emph{Observability:} \texttt{exp\_01\_optical\_axis\_isotropy} ($129^3$, both engines) -- isotropic pulse flattens perpendicular to $(1,1,-1)$ (ratio $\to 0.01$), flat-axis alignment $=1.0000$, $\omega$-independent, survives exact A=1 renormalization.
    & \texttt{PASS} \\

% ── Gauge sector: birefringence verdict ─────────────────────────────
Gauge-sector vacuum birefringence \emph{cancels} (null polarization split)
    & Induced Maxwell action $\tfrac12(\mathbf{E}^T\!\boldsymbol\varepsilon\,\mathbf{E} + \mathbf{B}^T\boldsymbol\mu^{-1}\mathbf{B})$ with $\boldsymbol\varepsilon = P = \{1,4,4\}$ (axis $(1,1,-1)$ suppressed) and $\boldsymbol\mu^{-1} = Q_B = \operatorname{adj}(\boldsymbol\varepsilon) = \{4,4,16\}$ (axis enhanced). The adjugate relation makes the photon dispersion a \emph{double} transverse root for every direction, so the two polarizations share $\omega^2 = k^T\!\boldsymbol\varepsilon\,k$ and the split vanishes
    & Isotropic-impedance product $\varepsilon_i\,(\mu^{-1})_i = \det\boldsymbol\varepsilon$ on every axis; the cancellation is prefactor-independent (immune to the deferred $1/g^2$) and holds for any hop vectors (Paper~VIII adjugate theorem)
    & \emph{Blocks from the engine geometry} (\texttt{exp\_03\_geometric\_blocks}): $\boldsymbol\mu^{-1}=\{4,4,16\}$ via the spatial plaquette holonomy, $\boldsymbol\varepsilon=\{1,4,4\}$ via the temporal plaquette holonomy off \texttt{HopOperator.step} (coupling fidelity $4\times10^{-19}$; $\omega$-cancellation $7\times10^{-15}$); adjugate $\boldsymbol\mu^{-1}=\operatorname{adj}(\boldsymbol\varepsilon)$ confirmed from the two engine blocks ($5\times10^{-15}$). \emph{Verdict:} the dispersion solver on the engine-extracted blocks gives $\max|v_+-v_-| = 1.6\times10^{-15}$ over 2000 directions (a detuned control gives $0.5$), and the anisotropy \emph{survives A=1 renormalization} in the multi-tick dynamics (correct senses). Companion of Paper~VIII (analytic derivation). \emph{Open item (stated):} the single-object dynamical (\texttt{Tr}\,$\ln T$ / orbital-susceptibility) extraction of the tensor -- a naive band sum is Ward-divergent on the flat optical axis and yields the wrong object; deferred to follow-on work.
    & \texttt{PASS} \\

% ── Gauge sector: common-mode speed anisotropy ──────────────────────
Common-mode (direction-dependent) vacuum speed, factor $\sim2$ about $(1,1,-1)$
    & Both transverse polarizations share $v^2(\hat{\mathbf{k}}) = \hat{\mathbf{k}}^T\!\boldsymbol\varepsilon\,\hat{\mathbf{k}} \in [1,4]$; the lattice has \emph{one fixed} hop-set (optical axis = the omitted 4th cube body-diagonal), so it is a single trigonal domain
    & The $O_h$ 4-domain average $\langle\boldsymbol\varepsilon\rangle = 3\mathbb{I}$, $\langle\boldsymbol\mu^{-1}\rangle = 8\mathbb{I}$ is isotropic, but it is a \emph{fictitious ensemble} the fixed lattice does not realize
    & \texttt{exp\_03\_geometric\_blocks}: the engine blocks give common-mode $v\in[1.00,2.00]$ (single-domain factor $2.00$); the $O_h$ average restores isotropy ($10^{-16}$). This is a \emph{speed} anisotropy, not birefringence, and is distinct from the (separately settled) null split. Whether the fixed lattice or an averaged ensemble is physical is the isotropy-restoration question shared with Paper~VIII.
    & \texttt{PART} \\

% ── Concordance ─────────────────────────────────────────────────────
Multi-channel concordance (P9)
    & Kinematic and gauge sectors inherit the \emph{same} optical axis $(1,1,-1)$ from the common bipartite geometry ($M_\text{eff}$ and $\mathbf{Q}$ special directions coincide)
    & N/A (cross-sector structural identity)
    & Axis-sharing is a \emph{structural} identity (derived): $M_\text{eff}$, $\boldsymbol\varepsilon$, and $\boldsymbol\mu^{-1}$ all inherit $(1,1,-1)$ from the same bipartite geometry. The kinematic dispersion is the one frontier-testable observational channel; the gauge sector's leading birefringence \emph{cancels} (null split, its own row), so the shared axis registers there as the direction of the common-mode speed anisotropy rather than a polarization split. P9 is a genuine cross-sector structural concordance; its strength as a multi-\emph{observable} empirical claim remains limited because the gauge birefringence is null and the gauge common-mode speed's observability is the open isotropy-restoration question.
    & \texttt{PART} \\

\hline
\end{longtable}
}
