% conclusion.tex
% No-go framing: the universal birefringence cancellation SURVIVES; the
% single-domain construction (bare matter + induced photon) is EXCLUDED as
% constructed. Both exclusions share the three-of-four root and an emergence /
% O_h-restoration escape.

The optical axis $\hat{\mathbf{n}}_* = (1,1,-1)/\sqrt3$, fixed geometrically as the
cube body-diagonal the hop set omits, organizes anisotropies in both the matter and the
gauge sector---and the honest account is a clean survivor beside a clean no-go.

\emph{What survives.} In the gauge sector the two geometric plaquette forms obey the
adjugate relation $\boldsymbol{\mu}^{-1} = \gamma\operatorname{adj}
(\boldsymbol{\varepsilon})$ (Theorem~\ref{thm:adjugate})---an identity for any hop
vectors, and a lattice instance of the classical impedance-matched (birefringence-free)
medium---which forces a doubled transverse root (Theorem~\ref{thm:doubled_root}): the
vacuum birefringence cancels. Given those geometric blocks this is \emph{universal}:
no three-hop lattice can be birefringent, independent of the particular hop set and of
every deferred normalization---though establishing the blocks as the full dynamical
induced action is a separate, still-open step (Sec.~\ref{sec:gauge_sector}). Read as a falsification test, the observed absence
of optical-axis vacuum birefringence is a stringent existing constraint this class of
models survives---a non-trivial postdiction, since the bounds predate the framework---and
the same datum excludes the CPT-odd, dimension-five and other birefringent
Lorentz-violating rivals while corroborating the framework's non-birefringent class. A
further non-detection no longer \emph{discriminates} the framework from standard
electrodynamics, so the forward frontier is dispersion, not birefringence.

\emph{What is excluded as constructed.} The lattice's actual single fixed hop set (one
trigonal domain) is a Lorentz violation in both sectors. The induced photon carries an
$O(1)$, dimension-four common-mode speed anisotropy (factor $\sim2$ about
$\hat{\mathbf{n}}_*$; three independent computations), excluded by isotropy experiments:
decisively for the geometric gauge completion, strong-provisionally for the dynamical
model pending the full effective-action Hessian. And the bare single-tick spinor, read
as literal laboratory matter, is excluded more sharply still
(Sec.~\ref{subsec:as_constructed}): its energy dispersion \emph{relative to the rest
phase} is the small dimension-six anisotropy of the sidereal channel, but its kinetic and
level structure---the quantity an isotropy experiment actually compares---is anisotropic
at order unity, its axial band is exactly flat (axial transport identically zero), and its
$\omega=0$ amplitude filter accumulates \emph{linearly} in the lattice spacing
($e$-folding times of days to minutes at Planckian $a$). The earlier expectation that
matter is spared because its anisotropy is dimension-six is therefore too glib: the
dimension-six number is the rest-relative energy only, and the observable matter sector is
$O(1)$ as constructed, just as the photon is.

Both exclusions have the one origin: the anisotropy is the single tensor
$\sum_a\mathbf{V}_a\mathbf{V}_a^{\!\top} = \{1,4,4\}$, which is $O(1)$ only because the hop
set omits the fourth cube diagonal $(1,1,-1)$; the four-diagonal set is exactly isotropic.
The optical axis this paper studies and both single-domain exclusions are the same fact
seen three times. Their escapes are the ones the program already owns and are plausibly
two faces of one extension: a vacuum that superposes the four diamond orientations so the
induced gauge loop sees the isotropic average, and the token$\to$matter emergence map
(with the minimum-probability floor $\delta p_\text{min}$) so that laboratory matter is an
emergent excitation rather than the bare spinor. The averaging is not new to this paper:
it is the same $O_h$ average by which Paper~II~\cite{menendez2026smderivation} recovers
isotropic Maxwell and by which the program obtains emergent Lorentz invariance
itself~\cite{menendez2026geometryfirst}, so what the single-domain construction exposes is
the observable cost of a foundational assumption the series already relies on. Whether an
$O_h$-restoring vacuum and an emergence map preserve the A=1 and Dirac structure is a
foundational question for future work; we state the no-go plainly rather than resolve it
here.

None of this touches what stands independently: the universal birefringence cancellation,
the adjugate theorem, and the exact closed-form substrate kinematics. The paper's result
is that the same geometry which makes discrete three-hop electromagnetism incapable of
birefringence also, on its single-domain face, excludes both its photon and its bare
matter as Lorentz-violating---and that saying so is what the framework's own audit
discipline requires.
