% falsifiability.tex -- the multi-channel falsifiability map.
%
% Author decision (handoffs 2026-07-18-paper04-falsifiability-do-both-decision +
% -birefringence-null-is-a-passed-falsifier): DO BOTH channels. 2.b (kinematic) is
% concrete here; 2.a (gauge photon) is a contingent OPEN follow-on. Reframe the
% birefringence null as a PASSED falsifiable prediction (not "not a falsifier"),
% with two honesty guardrails. SME placement verified against Kostelecky--Mewes.
%
% Backs the falsifiability audit row (Table~\ref{tab:audit}).

\label{sec:falsifiability}

The program's motivation for this paper is falsifiability, and the optical axis is a
parameter-free target. It is worth being precise about which of the framework's
predictions actually test it, because the most common objection gets the logic
backwards.

\subsection{The birefringence null is a passed test, not an absence of content}
\label{subsec:passed_null}

Discrete-spacetime proposals are often dismissed on the grounds that ``there is no
evidence of birefringence.'' A falsifiable prediction, however, is precisely one that
\emph{forbids} an observable outcome. The framework forbids optical-axis vacuum
birefringence---the doubled transverse root of Theorem~\ref{thm:doubled_root}---and
vacuum birefringence is both directly observable and among the most tightly bounded
quantities in physics. Its non-observation is therefore not silence; it is a stringent,
independently-measured constraint the framework \emph{survives} by construction---a
falsification test (an observed optical-axis birefringence would kill the framework) that
it passes. Because the bounds predate the framework, this is a postdiction rather than a
novel prediction; but it is a non-trivial one, and the distinction below is what makes it
worth stating.

The test moreover \emph{discriminates}. In the Standard-Model Extension (SME)
classification of Lorentz-violating photon
operators~\cite{kosteleckymewes2009,kosteleckyrussell2011}, operators of odd mass
dimension are CPT-odd and \emph{birefringent}; the leading Planck-suppressed such
operator is dimension five, giving a helicity-dependent vacuum birefringence. Many
Lorentz-violating models---in particular the CPT-odd, dimension-five and other
birefringent classes---populate this sector, and it is exactly the astrophysical
non-observation of vacuum birefringence that \emph{excludes} them. (The SME also contains
non-birefringent, dispersive, and vacuum-orthogonal classes; the claim is discrimination
against the \emph{birefringent} ones, not against all Lorentz violation.) This framework instead sits in the CPT-even, non-birefringent
sector: the induced photon action's adjugate closure (Theorem~\ref{thm:adjugate})
cancels the polarization split, placing it with the even-dimension, dispersive-but-not-
birefringent operators (the $c^{(6)}_{(I)jm}$ family). The same datum that refutes the
birefringent rivals therefore \emph{corroborates} this one. The null is not luck; it is
what the framework's operator class requires.

Two guardrails keep this honest. \emph{(a) Retrospectively}, the null is a passed
falsification test---a postdiction, since the bounds predate the framework---and a strong
one, because the natural birefringent alternatives fail the same test. \emph{(b) Prospectively}, a further non-detection of birefringence no longer
\emph{discriminates} this framework from standard electrodynamics---so the
forward-testable frontier is the dispersion channels below, not birefringence. And the
cancellation is a theorem \emph{conditional} on the constitutive blocks
(Sec.~\ref{sec:gauge_sector}); their derivation as the full dynamical vacuum response
is itself an open item (Sec.~\ref{subsec:gauge_channel}). The statement is thus
corroborated \emph{and} conditional---no more, and no less.

\subsection{The falsifiable content is dispersion, and it splits by field}
\label{subsec:falsifiability_map}

Because birefringence is forbidden, the forward-testable content is the
direction/energy-dependent dispersion, and the $\omega=0$ analysis of
Sec.~\ref{subsec:a1_normalization} forces it to split into distinct channels living in
distinct fields. We map them honestly, per channel:

{\small
\begin{center}
\begin{tabular}{@{}p{3.5cm}p{1.9cm}p{3.6cm}p{3.4cm}@{}}
\hline
\textbf{Channel} & \textbf{Field} & \textbf{Effect} & \textbf{Status} \\
\hline
Optical-axis birefringence & gauge photon & polarization split & \emph{Forbidden}---passed falsification test (postdiction; corroborated; conditional on the blocks) \\
Kinematic massive dispersion & spinor (matter) & dim-6 directional dispersion; sidereal & Real as \emph{substrate} kinematics; lab connection needs the emergence map (\texttt{PART}) \\
Kinematic massless / kinetic anisotropy & spinor (as-constructed) & $O(1)$ kinetic anisotropy + exact flat band + linearly-accumulating $\omega=0$ filter & \emph{Excluded as constructed}; open via the emergence map / $\delta p_\text{min}$ (\texttt{FAIL} as-constructed) \\
Gauge-sector photon speed & U(1) gauge & common-mode speed anisotropy & \emph{Tension}---$O(1)$ dim-4, excluded as a single domain; structural fix (four-axis / $O_h$) open (\texttt{FAIL} as-constructed) \\
\hline
\end{tabular}
\end{center}
}

\subsection{The kinematic channel (matter sector)}
\label{subsec:kinematic_channel_falsif}

The concrete falsifiable content of this paper is the massive-sector directional
dispersion of Sec.~\ref{sec:kinematic_channel} ($\omega\neq0$). Its low-energy form
$|\tilde H_\text{RGB}|^2 = 1 - \tfrac43|\mathbf{k}_\perp a|^2$ is a
\emph{dimension-six}-scale anisotropy: the fractional effect is $\sim(E a/\hbar c)^2
\sim (E/E_\text{Planck})^2$ if the lattice spacing is near Planckian. Two things must
be said plainly about its reach.

First, the \emph{magnitude} is small and the honest expectation is a bound, not a
detection. Dimension-six suppression is the most suppressed Lorentz-violating case. To
gauge how weak that makes the reach, take the generic assumption---common to
Planck-suppressed models, but \emph{not} a delivered result of this framework's own
photon channel---which, as constructed, carries an $O(1)$ tension of its own rather than
a suppressed effect (Sec.~\ref{subsec:gauge_channel})---that a
dimension-six effect of order $(a/\hbar c)^2$ exists: even then, the strongest current
photon dimension-six time-of-flight bound, LHAASO's observation of
GRB~221009A~\cite{lhaaso2024grb}, would only require $a \lesssim 3\times10^{-28}$~m
($\sim\!10^{7}\,\ell_\text{Planck}$). It is an external scale-setting figure for how far
above the Planck length a bound would sit, not a constraint this framework's photon
sector has itself established. The massive-sector channel is weaker still, forgoing the
cosmological baseline that gives photons their lever arm. A Planck-scale lattice sits
some twenty orders below current reach; the correctly scoped claim is a limit far above
the Planck scale, not a signal.

Second, and more usefully, the distinctive handle is the \emph{signature}, not the
magnitude. A lattice with a fixed spatial axis $\hat{\mathbf{n}}_*$ is anisotropic in a
way no isotropic theory reproduces: as the Earth rotates, the laboratory frame sweeps
through $\hat{\mathbf{n}}_*$, so the predicted effect \emph{modulates at the sidereal
frequency}. In SME language this is the statement that the anisotropic coefficients
($j\neq0$) are non-zero with a fixed, parameter-free angular pattern about
$\hat{\mathbf{n}}_*$~\cite{kosteleckymewes2009}. A sidereal modulation with this
specific pattern is a qualitatively distinctive target that an isotropic
Lorentz-violating model cannot fake, even where the amplitude is small. The natural
observational handles, stated with their caveats rather than as claims, are then: the
sidereal modulation of high-energy propagation; high-energy astrophysical
\emph{neutrinos} as the one massive channel that keeps a cosmological baseline; and the
$\omega=0$ magnitude anisotropy---a direction-dependent \emph{filtering} of amplitudes
(not an energy dispersion), the anisotropic spreading that \texttt{exp\_01} measures.

\paragraph{The as-constructed caveat.} These handles describe the \emph{substrate}
kinematics of the bare single-tick spinor. Read as a literal laboratory particle, that
spinor is excluded as constructed (Sec.~\ref{subsec:as_constructed}): its kinetic and
level structure is $O(1)$ anisotropic, its axial band is exactly flat, and its $\omega=0$
amplitude filter accumulates \emph{linearly} in $a$---$e$-folding times of days to minutes
at Planckian $a$, verified in \texttt{exp\_05\_matter\_sector\_order}, tens of orders above
co-magnetometer and matter-interferometer bounds. The sidereal massive-dispersion channel
is therefore a prediction of the \emph{substrate}, connected to an observable particle only
once the token$\to$matter emergence map---laboratory matter as an emergent, composite
session---is exhibited; until then the matter channels read literally are excluded, with the
emergence map (and the $\delta p_\text{min}$ floor that regulates the filter) the open route
to viability.

\subsection{The gauge-sector photon channel: a stated tension}
\label{subsec:gauge_channel}

The genuinely \emph{photonic} dispersion is \emph{not} delivered by the $\omega\to0$
limit of the spinor propagator, which does not disperse (Prop.~\ref{prop:massless_flat}).
On this lattice the photon is the induced gauge field, and the corresponding effect is
the gauge-sector \emph{common-mode} speed anisotropy $\omega^2 =
\gamma\,\mathbf{k}^{\!\top}\boldsymbol{\varepsilon}\,\mathbf{k}$
(Sec.~\ref{subsec:common_mode}), with $\boldsymbol{\varepsilon}$ eigenvalues
$\{1,4,4\}$---a factor-$\sim2$ direction-dependent vacuum speed of light whose ratio is
independent of the overall normalization $\gamma$. Honesty requires reporting where this leads,
because it is the one place the framework is presently in tension with observation.

\paragraph{The anisotropy is order unity, not suppressed.} Three independent routes
agree that the common-mode anisotropy is $O(1)$ and \emph{dimension-four} (present at
leading order, surviving the continuum limit), not a suppressed dimension-six effect:
the geometric induced blocks give a $\mathbf{k}$-independent factor-$\sim2$ speed
($\boldsymbol{\varepsilon}=\{1,4,4\}$); a Ward-safe physical-band vacuum-polarization
sweep gives an anisotropic (traceless) part that is $\sim\!50\%$ of the isotropic part
and stays constant as the photon wavevector $\to0$; and the static density response
(\texttt{exp\_03a}) is axis-anisotropic at the same order. (See
\texttt{notes/gauge\_common\_mode\_order.md}.)

\paragraph{As constructed, this is excluded.} An $O(1)$, dimension-four,
direction-dependent speed of light is a \emph{marginal} Lorentz violation---the
CPT-even, non-birefringent anisotropic-$k_F$ sector of the
SME~\cite{kosteleckymewes2009,kosteleckyrussell2011}, bounded by optical-cavity and
Michelson--Morley isotropy experiments to $\sim\!10^{-18}$. A factor-$\sim2$ effect,
sidereally modulated by the fixed axis, would have been seen long ago. Crucially, the
exclusion does not hinge on any cancellation between sectors. The induced photon's $O(1)$
anisotropy is bounded directly by optical-cavity and Michelson--Morley isotropy tests; and
the bare matter sector, far from providing an isotropic reference that could absorb the
photon's tilt as a coordinate choice, is \emph{itself} $O(1)$ anisotropic---an exact axial
flat band and an order-unity kinetic/level-structure anisotropy
(Sec.~\ref{subsec:as_constructed})---so it too is independently excluded. For the physical
vacuum taken as the single fixed trigonal domain of the A=1 hop set, both the induced
electromagnetism and the bare matter are therefore Lorentz-violating at order unity, and
\emph{excluded}.

\paragraph{The level of the claim.} We state the verdict at two strengths, because the
geometric and the fully dynamical blocks are established to different degrees. For the
\emph{geometric} single-domain gauge completion---the plaquette forms
$\boldsymbol{\varepsilon}=\sum_a\mathbf{V}_a\mathbf{V}_a^{\!\top}$ taken as the
constitutive tensors---the falsification is \emph{decisive}: the tensor is anisotropic by
construction and the exclusion is immediate. For the \emph{dynamical} single-domain A=1
model, the reading off \texttt{HopOperator.step} establishes the field--matter vertex but
not the full effective-action Hessian
$\varepsilon_{ij}\propto\partial^2\Gamma/\partial E_i\,\partial E_j$
(Remark~\ref{rem:vertex_not_action}), so the verdict is a \emph{strong provisional}
falsification: the three routes above (including a Ward-safe band calculation whose
anisotropy is finite as the photon momentum $\to0$) show that the physical response
contains an unsuppressed dimension-four anisotropic component, but the complete tensor
assembly (the orbital-susceptibility $\operatorname{Tr}\ln T$ object) remains open. The
``provisional'' qualifier is removed once that calculation confirms a nonzero
dimension-four traceless part; it need not reproduce $\{1,4,4\}$ exactly, as any
unsuppressed anisotropy remotely of order unity is already fatal.

\paragraph{Where the two sectors diverge---and where they do not.} It is tempting to say
the matter sector is protected: the matter dispersion's leading term is a \emph{scalar}
($|\sum_a e^{i\mathbf{k}\cdot\mathbf{V}_a}|^2$, isotropic at $\mathbf{k}=0$), so its
\emph{energy} anisotropy relative to the rest phase is pushed to the $O(k^2)$,
dimension-six correction, whereas the induced photon block is a \emph{tensor} from the
outset, $\boldsymbol{\varepsilon}=\sum_a\mathbf{V}_a\mathbf{V}_a^{\!\top}$, with its
anisotropy at $O(1)$, dimension-four. That contrast is real but only for the
energy-dispersion observable; it does \emph{not} make matter safe. As
Sec.~\ref{subsec:as_constructed} shows, the same operator's kinetic and level
structure---the quantity an isotropy experiment actually compares---is anisotropic at
order unity, and its axial band is exactly flat, so the bare matter sector is excluded as
constructed just as the photon is. The honest statement is thus not that Lorentz
invariance descends to matter but not the gauge field; it is that emergent leading-order
Lorentz invariance fails to descend to \emph{either} single-domain sector at the level of
observable anisotropy---a recognized naturalness difficulty of emergent-spacetime
constructions~\cite{collins2004}. And the whole effect is the single tensor
$\sum_a\mathbf{V}_a\mathbf{V}_a^{\!\top}$: it is anisotropic ($\{1,4,4\}$) precisely because
the hop set is \emph{three} of the four cube body-diagonals; the \emph{four}-diagonal set
gives $\sum_a\mathbf{V}_a\mathbf{V}_a^{\!\top}=4\,\mathbb{I}$, exactly isotropic. The
optical axis (this paper's subject) and the gauge-sector tension therefore share one
origin: the omission of the fourth diagonal $(1,1,-1)$---equivalently, the selection of a
single trigonal domain from the four.

\paragraph{The resolution is structural, and open.} Because the anisotropy is fixed by the
choice of hop set, the framework already admits constructions that remove it: including the
fourth axis (and, more generally, higher-dimensional and polygonal axis topologies the
underlying formalism permits) restores the cubic group. In particular, a construction in
which the vacuum superposes the four diamond orientations would present the induced gauge
action---a loop over the whole vacuum---with the $O_h$-symmetric average
($\langle\boldsymbol{\varepsilon}\rangle\propto\mathbb{I}$, isotropic, tension removed),
while laboratory matter, as an \emph{emergent} session rather than the bare spinor
(Sec.~\ref{subsec:as_constructed}), could carry its own viable dispersion. Two cautions
keep this honest, both raised by the same logic that forbids the naive $O_h$ average. The
superposition must be \emph{coherent}, not a classical ensemble: a statistical mixture of
orientations leaves every realized branch with its excluded $O(1)$ anisotropy. And
axis-bearing matter can \emph{decohere} that superposition, collapsing the induced medium
into a single excluded branch, so the sector decoupling must be \emph{exhibited, not
assumed}---it is dynamical hop-set geometry (orientation promoted to a quantum degree of
freedom), a genuine extension of the formalism. Whether such a construction, together with
the token$\to$matter emergence map, preserves the A=1 conservation and the Dirac/chiral
structure that the three-axis choice was made to secure is a foundational question we do
not settle here; it is the natural next research direction, and the honest status of both
single-domain sectors is that of an exclusion with a known-but-unbuilt structural
resolution, not a delivered prediction.

\paragraph{Scope.} This tension is specific to the gauge-sector common-mode speed. It does
\emph{not} bear on the birefringence null (which passes, Theorem~\ref{thm:doubled_root}),
the adjugate theorem, the kinematic directional dispersion, or results elsewhere in the
program; those stand independently of how the gauge-sector construction is resolved.

Taken together, the framework's falsifiability map is a clean survivor beside a clean
no-go. What survives: a forbidden birefringence the data have already corroborated
(excluding the birefringent rivals), and universally so---no three-hop lattice can produce
it. What is excluded as the single-domain construction stands: the induced-photon speed
anisotropy ($O(1)$, dimension-four), \emph{and} the bare matter sector itself (an exact
flat band, $O(1)$ kinetic anisotropy, and a linearly-accumulating filter), both pointing to
the same structural revision---an $O_h$-restoring vacuum and a token$\to$matter emergence
map. The substrate massive dispersion, with its distinctive sidereal signature, is a real
prediction awaiting that emergence map to reach a laboratory particle. The value is not one
frontier number but a candid, per-channel account of what tests the theory---including,
squarely, where it is presently tested and found wanting.
