% gauge_sector.tex -- the gauge-sector channel (merged from Paper VIII's
% electric-block derivation + Paper IV's from-engine verification).
%
% Load-bearing matrices are \input from sections/generated/*.tex, emitted by
% src/utilities/electric_induced_action.py via sympy.latex() (provably identical
% to the verifier). Do not hand-edit those fragments.
%
% Backs the gauge-sector audit rows (Table~\ref{tab:audit}); verified by
% src/utilities/electric_induced_action.py (symbolic), exp_04_electric_
% permittivity_extraction.py (engine read-off of eps), exp_03_geometric_blocks.py
% (both blocks + verdict), exp_03_dispersion_core.py (Fresnel solver).
%
% Review-corrected 2026-07-17: the plaquette forms are GEOMETRIC CANDIDATE
% response blocks (bare geometric plaquette actions), NOT a proven dynamical
% induced action; det(eps) prefactor (not 16); proportional closure gamma*adj(eps);
% exact adjugate does NOT survive O_h averaging; "constitutive closure" not
% "covariant completion"; the word "birefringence" is reserved for this section.

\label{sec:gauge_sector}

Where the kinematic channel (Sec.~\ref{sec:kinematic_channel}) lives in the matter
propagator, the gauge channel lives in the vacuum's response to a background
electromagnetic field. It is here, and only here, that the paper speaks of
\emph{birefringence} in its literal optical sense: two transverse polarizations
of one propagation vector, and whether their speeds split. The lattice couples the
two static fields on different footings. A magnetic field enters as a \emph{spatial}
link phase---the hop along a lattice direction $\mathbf{V}_a$ picks up the Peierls
phase $\exp(i\,\mathbf{A}\!\cdot\!\mathbf{V}_a)$, so a closed spatial plaquette
carries a gauge-invariant Wilson holonomy. An electric field enters instead as an
\emph{on-site} phase advance $\delta\phi(\mathbf{x}) = \omega + V(\mathbf{x})$, the
temporal potential $A_0$ added to the rest phase, with no spatial link. From these
two couplings we build two quadratic field forms, show they are bound by an exact
algebraic identity, and read off its consequence for polarization.

A caveat frames everything below. The two forms are the natural \emph{geometric}
(plaquette) quadratic forms the hop set supplies; we call them a permittivity
$\boldsymbol{\varepsilon}$ and an inverse permeability $\boldsymbol{\mu}^{-1}$
because they play those roles in the constitutive relation, but establishing them
as the \emph{dynamical} induced action---the second-order response obtained by
integrating the matter tokens out---is a separate step we do not complete here (it
is the stated open item of Sec.~\ref{subsec:gauge_scope}). What is rigorous is the
algebra of the two geometric forms and the polarization verdict it forces; that
verdict is prefactor-independent, so it survives whatever fixes the blocks' overall
normalization.

\subsection{The magnetic form (anchor)}
\label{subsec:magnetic_anchor}

A spatial plaquette spanned by two hop directions $\mathbf{V}_a,\mathbf{V}_b$ has,
for a uniform field, holonomy $(\mathbf{V}_a\!\times\!\mathbf{V}_b)\!\cdot\!\mathbf{B}$;
summing the squared holonomy over the three independent bipartite plaquettes gives
the $O(B^2)$ form
$\text{density}_B = \sum_{a<b}\bigl((\mathbf{V}_a\!\times\!\mathbf{V}_b)\!\cdot\!
\mathbf{B}\bigr)^2 = \mathbf{B}^{\!\top} Q_B\,\mathbf{B}$, with
\begin{equation}
    \label{eq:magnetic_Q}
    \input{sections/generated/magnetic_Q}
\end{equation}
whose eigenvalues are $\{4,4,16\}$: the optical axis $(1,1,-1)$ is \emph{enhanced}
(eigenvalue $16$), the perpendicular plane doubly degenerate at $4$. This is the
Paper~I Appendix~B tensor~\cite{menendez2026geometryfirst}, reproduced from the
engine's Peierls links by \texttt{dcl-core}'s \texttt{exp\_04}~\cite{dclcore} and
again here (\texttt{exp\_03\_geometric\_blocks}) to machine precision; it is the
exact consistency anchor throughout (Table~\ref{tab:audit}).

\subsection{The electric form}
\label{subsec:electric_block}

The on-site advance $\delta\phi = \omega + V(\mathbf{x})$ is itself a \emph{temporal}
link---a site joined to itself at the next tick---so a temporal plaquette spanned by
one hop direction $\mathbf{V}_a$ and the tick direction closes on the lattice. For a
uniform static field $V(\mathbf{x}) = -\mathbf{E}\!\cdot\!\mathbf{x}$ its holonomy is
the on-site phase difference along the hop,
\begin{equation}
    \label{eq:temporal_holonomy}
    \Theta_a(\mathbf{x}) = \delta\phi(\mathbf{x}) - \delta\phi(\mathbf{x}+\mathbf{V}_a)
        = V(\mathbf{x}) - V(\mathbf{x}+\mathbf{V}_a)
        = (\mathbf{V}_a\!\cdot\!\mathbf{E})\,a_t,
\end{equation}
the rest phase cancelling around the loop ($a_t$ is the tick's temporal extent).
Where the magnetic loop needs two spatial directions and a cross product, the
electric loop needs one. Summing the squared holonomy over the three hop directions
gives $\text{density}_E = \sum_a\Theta_a^2 = a_t^2\,\mathbf{E}^{\!\top}
\boldsymbol{\varepsilon}\,\mathbf{E}$, with
\begin{equation}
    \label{eq:electric_P}
    \input{sections/generated/electric_P}
\end{equation}
eigenvalues $\{1,4,4\}$: the same axis is now \emph{suppressed} (eigenvalue $1$),
the perpendicular plane again at $4$. The electric form is the mirror of the
magnetic one---axis suppressed where the magnetic axis is enhanced---and its sense
matches Paper~IV's static screen \texttt{exp\_03a}.

This form is read off the engine's actual tick evolution, not merely posited.
Seeding a uniform real probe state and taking one step of the hop operator, the
on-site term $i\sin(\delta\phi/2)\,\psi_R$ is purely imaginary while the kinetic hop
is purely real, so $\operatorname{Im}(\psi_R^{\text{new}})/\psi_R = \sin(\delta\phi/2)$
recovers $\delta\phi = \omega + V(\mathbf{x})$ from the step output (to $10^{-19}$);
$\boldsymbol{\varepsilon}$ is extracted from that recovered phase
(\texttt{exp\_04\_electric\_permittivity\_extraction}), with the tick weight
$a_t=1$ measured, not assumed.

\begin{remark}[What the read-off establishes, and what it does not]
\label{rem:vertex_not_action}
Recovering $\delta\phi = \omega + V$ from one \texttt{HopOperator.step} validates the
field--matter \emph{vertex} (the coupling in the tick rule), not the quadratic
effective action $\Gamma^{(2)}[\mathbf{E}]$. A general $D_{3d}$-invariant quadratic
form has \emph{two} independent coefficients (an axial and a transverse one), so the
$1{:}4$ eigenvalue ratio is a property of the equal-weight single-hop plaquette sum,
\emph{not} something $D_{3d}$ symmetry forces; cross-plaquette terms could reweight
it. To promote $\boldsymbol{\varepsilon}$ from ``geometric candidate for the electric
response block'' to ``the induced permittivity'' one needs the effective-action
Hessian $\varepsilon_{ij} \propto \partial^2\Gamma[\mathbf{A}]/\partial E_i\,\partial
E_j\big|_{\mathbf{A}=0}$ with $\Gamma[\mathbf{A}] = -\log\det T[\mathbf{A}]$, or a
directly measured second-order response. That step is deferred
(Sec.~\ref{subsec:gauge_scope}); the same caveat touches the magnetic form
(Eq.~\eqref{eq:magnetic_Q}), which is a prescribed geometric plaquette action rather
than an effective-action Hessian.
\end{remark}

\subsection{The constitutive closure}
\label{subsec:constitutive_closure}

The two geometric forms are not independent. Written through the hop-vector matrix
$M = [\mathbf{V}_1\ \mathbf{V}_2\ \mathbf{V}_3]$, the electric form is the Gram
matrix $\boldsymbol{\varepsilon} = \sum_a\mathbf{V}_a\mathbf{V}_a^{\!\top} =
MM^{\!\top}$, and the magnetic form is the Gram matrix of the pairwise cross
products, which is exactly its adjugate.

\begin{theorem}[Adjugate relation]
\label{thm:adjugate}
For any three spatial hop vectors, the magnetic form is the adjugate of the electric
one, $Q_B = \operatorname{adj}(\boldsymbol{\varepsilon})$:
\begin{equation}
    \label{eq:adjugate}
    \input{sections/generated/adjugate_relation}
\end{equation}
\end{theorem}
\begin{proof}
The cross products $\mathbf{V}_a\!\times\!\mathbf{V}_b$ are the rows of
$\operatorname{adj}(M)$, so $Q_B = \operatorname{adj}(M)^{\!\top}\operatorname{adj}(M)
= \operatorname{adj}(M^{\!\top})\operatorname{adj}(M) = \operatorname{adj}(MM^{\!\top})
= \operatorname{adj}(\boldsymbol{\varepsilon})$. Verified symbolically for general
vector components in \texttt{src/utilities/electric\_induced\_action.py}.
\end{proof}

Theorem~\ref{thm:adjugate} is general---it holds for any three hop vectors, not just
the octahedral ones. The A=1 lattice is the special case:

\begin{corollary}[The lattice blocks]
\label{cor:lattice_blocks}
For the hop set $\mathbf{V}_1=(1,1,1)$, $\mathbf{V}_2=(1,-1,-1)$,
$\mathbf{V}_3=(-1,1,-1)$, the electric form of Eq.~\eqref{eq:electric_P} has
eigenvalues $\{1,4,4\}$ and the magnetic form of Eq.~\eqref{eq:magnetic_Q} has
eigenvalues $\{16,4,4\} = \operatorname{adj}$, uniaxial about $(1,1,-1)$ in opposite
senses (axis suppressed for $\boldsymbol{\varepsilon}$, enhanced for $Q_B$).
\end{corollary}

The physically relevant statement is the \emph{proportional} closure
$\boldsymbol{\mu}^{-1} = \gamma\,\operatorname{adj}(\boldsymbol{\varepsilon})$ for a
positive scalar $\gamma$: the electric and magnetic sectors may carry different
overall normalizations (Sec.~\ref{subsec:gauge_scope}), and the exact equality
$\gamma=1$ is the special case realized by the raw lattice forms. Either way, the
adjugate carries each eigenvalue of $\boldsymbol{\varepsilon}$ to the product of the
other two, so every principal direction obeys $\varepsilon_i\,(\mu^{-1})_i =
\gamma\det\boldsymbol{\varepsilon}$: the impedance product is \emph{isotropic} even
though $\boldsymbol{\varepsilon}$ and $\boldsymbol{\mu}^{-1}$ are each anisotropic,
in opposite senses about $(1,1,-1)$. For the lattice values,
$\varepsilon_a\mu^{-1}_a = 1\cdot16 = 16 = 4\cdot4 = \varepsilon_p\mu^{-1}_p$.

\subsection{The birefringence verdict}
\label{subsec:birefringence}

Isotropic impedance is exactly the condition that removes vacuum birefringence, and
it upgrades to a theorem.

\begin{theorem}[Doubled transverse root; no gauge-sector birefringence]
\label{thm:doubled_root}
For a plane wave of unit direction $\hat{\mathbf{k}}$ in a medium with
$\boldsymbol{\mu}^{-1} = \gamma\operatorname{adj}(\boldsymbol{\varepsilon})$, the
Fresnel determinant $\det\bigl(K\,\boldsymbol{\mu}^{-1}K + \omega^2
\boldsymbol{\varepsilon}\bigr)$, $K = [\hat{\mathbf{k}}]_\times$, factorizes---for a
\emph{general} symmetric $\boldsymbol{\varepsilon}$ and general direction, with no
shared eigenvectors required---as
\begin{equation}
    \label{eq:dispersion}
    \input{sections/generated/dispersion_factorization}
\end{equation}
The transverse root is \emph{doubled}, so both polarizations share the single
dispersion
\begin{equation}
    \label{eq:shared_dispersion}
    \input{sections/generated/shared_dispersion}
\end{equation}
and the polarization split vanishes identically,
\begin{equation}
    \label{eq:null_split}
    \input{sections/generated/birefringence_condition}
\end{equation}
Gauge-sector vacuum birefringence cancels.
\end{theorem}

The general prefactor is $\det\boldsymbol{\varepsilon}$ (the value $16$ is that
determinant for the lattice blocks, not a universal constant), and the $\omega^2=0$
root dropped above is the unphysical longitudinal mode. Because the split depends
only on
the anisotropy ratios---fixed by the geometry through Theorem~\ref{thm:adjugate}---and
not on the overall block scales, the cancellation is invariant under independent
rescaling of $\boldsymbol{\varepsilon}$ and $\boldsymbol{\mu}^{-1}$: it survives the
proportional closure for any $\gamma$, and holds for any hop vectors. This is the one
gauge-sector result that is unconditional given the blocks. An engine-level and a
purely numerical check corroborate it: \texttt{exp\_03\_geometric\_blocks} extracts
both blocks from the engine, confirms $\boldsymbol{\mu}^{-1} =
\operatorname{adj}(\boldsymbol{\varepsilon})$ from them to $5\times10^{-15}$, and the
Fresnel solver (\texttt{exp\_03\_dispersion\_core}) returns $\max|v_+-v_-| =
1.6\times10^{-15}$ over $2000$ directions, while a deliberately detuned control
(breaking the adjugate relation) returns a split of $0.5$. This is an independent
code-path verification of the theorem---a separate implementation, but within the
same framework and engine---not an independent physical measurement.

\subsection{The common-mode speed anisotropy}
\label{subsec:common_mode}

The null split does not make the sector isotropic. The shared dispersion
$\omega^2 = \gamma\,\mathbf{k}^{\!\top}\boldsymbol{\varepsilon}\,\mathbf{k}$ is
direction-dependent: the common phase speed $v^2(\hat{\mathbf{k}}) =
\gamma\,\hat{\mathbf{k}}^{\!\top}\boldsymbol{\varepsilon}\,\hat{\mathbf{k}}$ ranges over
$\gamma$ times the eigenvalues of $\boldsymbol{\varepsilon}$, $v^2\in[\gamma,4\gamma]$.
The overall scale $\gamma$ (the relative electric/magnetic normalization) is
undetermined, but the \emph{ratio} $v^2_{\max}/v^2_{\min} = 4$---a factor-$\sim2$
anisotropy of the vacuum speed about $(1,1,-1)$, shared by both polarizations---is
independent of it and cannot be tuned away. This is a \emph{common-mode} speed
anisotropy, not birefringence, and it is not small; the engine blocks give the ratio
$v_{\max}/v_{\min}=2.00$.

\begin{remark}[Single domain versus $O_h$ restoration]
\label{rem:oh_restoration}
The anisotropy is the response of a single trigonal domain. Averaging the two blocks
over the four ways of choosing three of the four cube body-diagonals as hop
directions restores full cubic symmetry,
\begin{equation}
    \label{eq:oh_average}
    \input{sections/generated/oh_average}
\end{equation}
so the $O_h$-restored medium is isotropic. Note, however, that the \emph{exact}
adjugate identity does \emph{not} survive this average: $\operatorname{adj}(3\mathbb{I})
= 9\mathbb{I} \neq 8\mathbb{I}$, so $\langle\boldsymbol{\mu}^{-1}\rangle =
\tfrac89\operatorname{adj}(\langle\boldsymbol{\varepsilon}\rangle)$---it is the
\emph{proportional} closure ($\gamma=\tfrac89$) that survives. The averaged medium is
non-birefringent because it is \emph{isotropic}, not because the exact adjugate
relation persists. The physical lattice, though, has one fixed hop set---its optical
axis is a specific body-diagonal---so the four-domain average is a fictitious ensemble
it does not realize, and there is no dynamical mechanism (for this construction) that
performs it. The physical A=1 lattice is thus the \emph{single-domain} case, and the
factor-$\sim2$ common-mode anisotropy is a real, $O(1)$, dimension-four effect (three
independent routes, Sec.~\ref{sec:falsifiability})---which, being a direction-dependent
speed of light, is \emph{excluded} by isotropy experiments as constructed. Only a
different, $O_h$-restoring construction (a four-orientation vacuum, or a more symmetric
hop set) would remove it, and whether such a construction preserves the A=1 and Dirac
structure is an open foundational question (Sec.~\ref{subsec:gauge_channel}). This
common-mode tension is entirely separate from the polarization verdict, which cancels in
\emph{every} domain regardless.
\end{remark}

\begin{remark}[The averaging is a foundational assumption inherited from Papers~I--II]
\label{rem:paper2_inheritance}
The four-domain average of Remark~\ref{rem:oh_restoration} is not a device peculiar to
the photon. Democratizing over the four choices of ``three of the four body-diagonals''
is arithmetically the same operation as the $O_h$ average $\langle Q\rangle =
(\operatorname{Tr}Q/3)\,\mathbb{I} = 8\mathbb{I}$ that Paper~II applies to the
\emph{identical} $\{4,4,16\}$ tensor to recover isotropic Maxwell and the gauge-coupling
ratios~\cite{menendez2026smderivation}, and---more fundamentally---it is the same
averaging by which the program obtains emergent Lorentz invariance itself ($SO(3,1)$ from
the $O_h$-averaged Dirac operator~\cite{menendez2026geometryfirst,menendez2026smderivation}).
Whether that average is \emph{physically realized}, as
opposed to a symmetry of an ensemble the single fixed vacuum does not sample, is therefore
a foundational question for the whole construction, not a local difficulty of the gauge
sector. The induced photon is simply the first place its cost is quantitative and
observable. (Even the order-$48$ automorphism group invoked by the average is the
signed-permutation group of the three hop-basis vectors, only twelve of whose elements are
orthogonal in Euclidean $\mathbb{R}^3$; the point symmetry a single physical domain
actually realizes is the trigonal $D_{3d}$.)

How the assumption plays out in the two sectors is subtler than a clean split. It is true
that the spinor's \emph{energy} dispersion relative to the rest phase is the
dimension-six, Planck-suppressed anisotropy of Sec.~\ref{sec:falsifiability}; but the bare
single-tick spinor's \emph{kinetic and level structure}---the quantity an isotropy
experiment actually compares---is anisotropic at order unity
(Sec.~\ref{subsec:as_constructed}), so read as a literal laboratory particle it is
excluded as constructed just as the induced photon is. Both sectors therefore pay the
single-domain cost, and both require a more complete construction to escape it---the
token$\to$matter emergence map for the spinor, an $O_h$-restoring (four-orientation)
vacuum for the gauge loop, plausibly two faces of the same extension. What they share at
the root is the arbitrary omission of the fourth diagonal that fixes the optical axis this
paper studies.
\end{remark}

The optical axis itself is geometric: as \emph{axes} the three hop directions are
three of the four cube body-diagonals, and $(1,1,-1)$ is the fourth---the one the hop
set omits. Selecting three of four diagonals breaks $O_h$ to the trigonal $D_{3d}$
about the omitted diagonal, and a $D_{3d}$-invariant quadratic form is uniaxial,
which is why both blocks come out $\{\text{axis};4,4\}$.

\subsection{Scope: structure, normalization, and the open dynamical tensor}
\label{subsec:gauge_scope}

Three items are deferred rather than concealed. First, \emph{normalization}: a single
common factor (a one-loop $1/g^2$) multiplying both parts of the action cancels from
the source-free equations of motion and so does \emph{not} set the photon speed; the
speed is set by the \emph{relative} electric/magnetic normalization---the tick-weight
and the temporal-versus-spatial lattice scales---which is the scalar $\gamma$ of the
proportional closure and is left undetermined. Because the birefringence verdict is
rescaling-invariant, this does not weaken it, but it means the sector fixes
anisotropic \emph{structure}, not the absolute vacuum speed.

Second, the \emph{dynamical} induced action. The blocks above are bare geometric
plaquette forms; a genuinely matter-integrated effective action is a loop-level
object, and demonstrating that integrating out the tokens reproduces
Eqs.~\eqref{eq:magnetic_Q}--\eqref{eq:electric_P}---with these weights and no
cross-plaquette terms---is the content of Remark~\ref{rem:vertex_not_action}. The
route is the second-order response of the physical band of the one-period transfer
operator $T(\mathbf{k})$ summed over the Brillouin zone ($\operatorname{Tr}\ln T$). We
do not carry it out and do not claim it. A naive band sum is Ward-divergent along the
optical axis (the physical band is exactly flat there, so the intraband denominator
$\lambda(\mathbf{k})-\lambda(\mathbf{k}\pm\mathbf{q})$ vanishes); once the
occupation-cancelled diamagnetic term is subtracted, the naive current--current object
returns the \emph{wrong} tensor (uniaxial about $(1,-1,0)$), and the correct assembly
is the orbital-susceptibility (Berry-curvature plus quantum-metric) combination. This
single-object dynamical extraction is a stated open item, deferred to follow-on work.
Since the geometric and effective-action routes must compute the same induced action,
and the geometric route already gives Eqs.~\eqref{eq:magnetic_Q}--\eqref{eq:electric_P},
finishing it is a matter of methodological completeness, not of deciding the verdict.

The status of the three claims is therefore distinct (Table~\ref{tab:audit}): the null
polarization split, \emph{conditional} on the blocks, is settled; the derivation of the
blocks as the actual dynamical effective action is partial; and the common-mode speed
anisotropy is a stated \emph{tension}---$O(1)$ and dimension-four, excluded as a single
domain (Sec.~\ref{sec:falsifiability}), with only its structural resolution open. Naming
the word once more, for precision: the sector predicts the \emph{absence} of gauge-sector
vacuum birefringence about $(1,1,-1)$; the accompanying common-mode speed anisotropy is,
for the single-domain construction, a real and excluded Lorentz violation, whose remedy
is a more symmetric construction of the hop set.
