% introduction.tex
% Orientation for the merged paper: the optical axis as the geometric spine; the
% two anisotropies (kinematic directional dispersion + gauge-sector birefringence
% cancellation); scope + honest status. The audit table is included in the Scope
% subsection so the reader meets it before any status-dependent claim.
% Terminology (external review, 2026-07-17): "birefringence" is reserved for the
% gauge-sector result; the kinematic effect is directional dispersion anisotropy.

\label{sec:introduction}

A crystal is optically anisotropic when its lattice singles out an axis. The A=1
discrete causal lattice of Paper~I~\cite{menendez2026geometryfirst} singles out an
axis too, but the ``crystal'' is spacetime itself: its bipartite hop set---the three
RGB vectors $\mathbf{V}_1=(1,1,1)$, $\mathbf{V}_2=(1,-1,-1)$,
$\mathbf{V}_3=(-1,1,-1)$ and their negatives---uses three of the four cube
body-diagonals and leaves the fourth, $\hat{\mathbf{n}}_* = (1,1,-1)/\sqrt3$, out.
That omission breaks the cubic point group $O_h$ to the trigonal $D_{3d}$ about
$\hat{\mathbf{n}}_*$ and fixes an optical axis geometrically, with no new postulate.
This paper works the resulting anisotropy to closed form in the two places it
appears---the matter propagator and the vacuum's gauge response---which share the
one axis.

The motivation is falsifiability. A geometry-fixed optical axis is a parameter-free
target, and the question the paper must answer honestly is which of its two
anisotropies could actually \emph{falsify} the framework. The answer is not the one
the word ``birefringence'' might suggest: the gauge-sector birefringence turns out to
\emph{cancel} (Sec.~\ref{sec:gauge_sector}), a consistency null rather than a frontier
test---so the DCL is, in this precise sense, \emph{not} falsifiable through
optical-axis birefringence. The candidate falsifiable content is instead the
directional dispersion (Sec.~\ref{sec:kinematic_channel}); as we show it is a
massive-sector effect, and its extension to the astrophysical \emph{photon} bounds
that would sharpen it into a frontier test is the paper's principal open item. Being
clear about which anisotropy tests the theory, which merely checks its consistency,
and what is not yet earned is the point of the exercise.

The two effects are physically different, and we name them differently. The
\emph{kinematic} anisotropy (Sec.~\ref{sec:kinematic_channel}) follows from Paper~I's
single-tick spinor transfer operator: its entire wavevector dependence passes through
one scalar structure factor whose magnitude is flat along $\hat{\mathbf{n}}_*$ and
curved perpendicular to it. This is a uniaxial \emph{directional dispersion
anisotropy}---a contrast between propagation \emph{directions}, maximal across the
axis and vanishing along it---rather than a polarization split at a fixed direction,
which is what ``birefringence'' denotes; we reserve that word for the gauge sector.
Because the transfer operator is non-unitary away from the axis, we are careful to
separate what its eigenphase magnitude establishes (an exact axial--transverse
anisotropy of the structure factor) from the further physical reading of it as a
propagating-mode group velocity, which we state as requiring the A=1 normalization
rather than assuming it.

The \emph{gauge} anisotropy (Sec.~\ref{sec:gauge_sector}) is the vacuum's response to
a background electromagnetic field, and here the outcome is the opposite of a new
signal. The lattice supplies two geometric plaquette quadratic forms---an electric
permittivity $\boldsymbol{\varepsilon}$ from the on-site temporal coupling and a
magnetic inverse permeability $\boldsymbol{\mu}^{-1}$ from the spatial Peierls
holonomy---and they obey the exact algebraic identity $\boldsymbol{\mu}^{-1} =
\operatorname{adj}(\boldsymbol{\varepsilon})$. This closure makes the two transverse
photon polarizations share a single dispersion, so the gauge-sector vacuum
\emph{birefringence cancels}; what remains is a direction-dependent common-mode speed
affecting both polarizations equally, whose observational status is open. The
adjugate theorem and its doubled-root consequence are the paper's central algebraic
result. We are equally explicit about the load-bearing caveat: the two forms are the
natural \emph{geometric} plaquette candidates for the response blocks, and
establishing them as the full \emph{dynamical} induced action---the second-order
effective action obtained by integrating out the matter---is a step we do not
complete, so the cancellation theorem is rigorous conditional on the proposed blocks.

This gauge-sector material was developed first as a separate companion study; it is
merged here because the shared optical axis is the natural unity of the two effects.
The gauge blocks are additionally read off the running lattice engine and the null
verdict re-derived through a separate code path, which we describe as an independent
code-path verification---a useful cross-check within the same framework, engine, and
geometric objects---rather than an independent physical measurement.

\subsection{Scope}
\label{subsec:scope}

Derived here and settled: the closed-form kinematic directional dispersion anisotropy
(Sec.~\ref{sec:kinematic_channel}); the adjugate constitutive closure and the null
gauge-sector birefringence \emph{conditional on the geometric blocks}
(Sec.~\ref{sec:gauge_sector}); and the structural axis-sharing of the two effects.
Explicitly deferred: the derivation of the gauge blocks as the actual dynamical
induced action (a matter-integrated effective action; the single-object
$\operatorname{Tr}\ln T$ tensor extraction is a stated open item); the relative
electric/magnetic normalization $\gamma$ that fixes the absolute vacuum speed (a
common coupling cancels and does not set it); and the observational status of the
common-mode speed anisotropy (the single-domain-versus-$O_h$-restoration question).
The physical reading of the kinematic eigenphase as a photon energy dispersion, and
the identification of the massless spinor mode with a photon, are likewise flagged
where they exceed what is proven. The audit table (Table~\ref{tab:audit}) is the
canonical record of which claims are \texttt{PASS}, \texttt{PART}, \texttt{STUB}, or
\texttt{FAIL}.

\input{sections/audit_table}
