% reproducibility.tex
\label{sec:reproducibility}

From a fresh clone at the release tag:

\paragraph{Environment.} Python (see \texttt{virtual-env-requirements.txt} for the
pinned interpreter and the \texttt{dcl\_core} engine version); the experiments use
only \texttt{numpy} and \texttt{sympy} for the closed-form checks, and
\texttt{dcl\_core} for the engine read-offs. Install the manifest into a virtual
environment rather than re-listing dependencies here.

\paragraph{Build the paper.} \texttt{make paper} (pdf\LaTeX\ + Bib\TeX) writes
\texttt{build/Paper.pdf}.

\paragraph{Verify the headline results.} Each experiment prints its own
\texttt{PASS}/\texttt{FAIL} summary:
\begin{itemize}
  \item \texttt{python src/experiments/exp\_02\_closed\_form\_split\_check.py} ---
    the closed-form kinematic identities (the \texttt{PASS} substrate content of the
    kinematic row).
  \item \texttt{python src/experiments/exp\_05\_matter\_sector\_order.py} --- the
    as-constructed matter exclusion (flat band, $O(1)$ anisotropy, linear filter).
  \item \texttt{python src/experiments/exp\_03\_geometric\_blocks.py} and
    \texttt{exp\_03\_dispersion\_core.py} --- the adjugate closure and the null
    polarization split.
  \item \texttt{python src/utilities/electric\_induced\_action.py} --- the symbolic
    derivation and (re)generation of the displayed equations.
\end{itemize}

\paragraph{Where to look next.} The audit table (Table~\ref{tab:audit}) maps every
claim to its backing script or note; Appendix~\ref{sec:band_sum} records the
Ward-safe band-sum estimate behind the dynamical common-mode verdict.
