% introduction.tex

\label{sec:introduction}

A discrete lattice that is meant to reproduce electromagnetism must, at low
energy, induce the Maxwell action---and it must do so for \emph{both} field
sectors on the same footing. On the A=1 Discrete Causal Lattice the magnetic
sector is settled: a background magnetic field enters the tick rule as a spatial
link phase (the Peierls substitution), a closed spatial plaquette carries a
gauge-invariant Wilson-loop holonomy, and its $O(B^2)$ induced action is the
tensor $Q_B$ with eigenvalues $\{4,4,16\}$ and optical axis $(1,1,-1)$ (Paper~I,
Appendix~B~\cite{PaperI}), reproduced from the engine's own link phases by
dcl-core \texttt{exp\_04}. The electric sector is not settled, and cannot be read
off the same way: an electric field enters the tick rule as an \emph{on-site},
mass-like phase advance $\delta\phi(x)=\omega+V(x)$, with no spatial link and so,
at first sight, no loop to extract a holonomy from. The corresponding electric
action block---the permittivity $\varepsilon$---is therefore absent from Paper~I
and from the Standard-Model derivation of Paper~II~\cite{PaperII}, and, because the
framework establishes only the spatial point group $O_h$ and not the Lorentz
boosts, there is no symmetry route that would fix $\varepsilon$ from the known
magnetic $\mu^{-1}$. The electric block must be derived directly; that is the task
of this paper.

The stake is concrete. The companion Paper~IV~\cite{PaperIV} asks whether the
lattice vacuum is birefringent about its optical axis $(1,1,-1)$---a binary,
near-falsifying question---and that verdict cannot be computed without the
covariant $(\varepsilon,\mu^{-1})$ pair. We supply it. We show that the on-site
phase advance is itself a \emph{temporal} link, so a temporal plaquette closes and
yields $\varepsilon=\sum_a V_a V_a^{\!\top}$ (eigenvalues $\{1,4,4\}$, the same
axis now \emph{suppressed}); that the magnetic block is the exact \emph{adjugate}
of the electric one, $\mu^{-1}=\operatorname{adj}(\varepsilon)$; and that this
adjugate relation forces the photon dispersion to a doubled transverse root, so
gauge-sector vacuum birefringence \emph{cancels}. The permittivity is not merely
posited: it is read directly off the engine's tick evolution, and the whole result
is confirmed independently by Paper~IV from an implementation of its own.

\subsection{Scope}
\label{subsec:scope}

This is a symbolic-derivation paper: every load-bearing claim is backed by a
runnable script whose printed output the audit table cites, and the derived
equations are generated from those verified expressions rather than transcribed.
In scope are the electric block, the covariant completion, and the birefringence
verdict. Two matters are deferred and stated plainly rather than hidden. First,
the shared photon speed is direction-dependent at the un-averaged (single-domain)
level---a \emph{common-mode} anisotropy that is not birefringence and whose
restoration to isotropy is a separate, undecided question. Second, the A=1 engine
carries only \emph{static} background fields; there is no dynamical gauge field and
hence no literal propagating photon to clock, so ``dispersion'' here means the
Fresnel analysis of the induced $(\varepsilon,\mu^{-1})$ tensors, and the
independent numerical content is the survival of the induced-action anisotropy
under the tick rule together with that dispersion verdict. The overall
one-loop coupling $1/g^2$ (the induced-action magnitude) is inherited as open from
Paper~I; it is immaterial to the birefringence verdict, which depends only on the
anisotropy \emph{ratios}.

Table~\ref{tab:audit} is the canonical record of which claims are
\texttt{PASS}/\texttt{PART}/\texttt{STUB}; the prose does not exceed it.

\input{sections/audit_table}
