\documentclass[12pt, letterpaper]{article}

%% Build from the paper/ directory:
%%   cd paper && pdflatex main.tex
%% or use the root build system:
%%   make paper          (from repo root)
%%   ./build.sh paper    (POSIX)
%%   build.cmd paper     (Windows cmd)

\input{macros/packages}
\input{macros/commands}

%% Page geometry (also set here as a fallback in case the package path fails)
\geometry{top=1in, bottom=1in, left=1in, right=1in}

%% ---------------------------------------------------------------
%% TITLE PAGE
%% ---------------------------------------------------------------
%% TODO: replace title, subtitle, author, ORCID, email.
%% TODO: at release time, fill in the \thanks{} block with version,
%%       Zenodo DOI, GitHub URL, and any series identifier.

\title{Optical-Axis Anisotropy on the A=1 Discrete Causal Lattice:\\
       Birefringence Cancellation and a Single-Domain No-Go%
       \thanks{Version 1.0.  Paper~IV of the A=1 Discrete Causal
       Lattice series; takes up follow-on item~\#14
       (\emph{Balanced Equations and Birefringent Channels}) of
       Paper~I's \texttt{notes/follow\_on\_implications.md} catalogue. This
       paper subsumes the former companion study of the electric induced-action
       block (Paper~VIII), merged here since the shared optical axis is the
       natural unity of the two effects.
       Builds on \emph{Geometry First}
       (Paper~I,
       \href{https://doi.org/10.5281/zenodo.20613677}{DOI:
       10.5281/zenodo.20613677}) and \emph{Geometry Forces Physics}
       (Paper~II,
       \href{https://doi.org/10.5281/zenodo.20292158}{DOI:
       10.5281/zenodo.20292158}).
       Paper~III (tidal ionization) is deferred pending the
       $\delta p_\text{min}$ investigation.
       \href{https://github.com/JackDMenendez/dcl-paper-04-optical-axis-birefringence}%
            {GitHub: JackDMenendez/dcl-paper-04-optical-axis-birefringence}.
       \href{https://doi.org/10.5281/zenodo.21435952}{DOI:
            10.5281/zenodo.21435952}.}}
\author{Jack Menendez\\
        \small Eugene, OR\\
        \small ORCID: \href{https://orcid.org/0009-0003-1166-307X}{0009-0003-1166-307X}\\
        \small \href{mailto:jack@geometryinducedphysics.org}{jack@geometryinducedphysics.org}\\
        \small Web: \href{https://geometryinducedphysics.org/}{geometryinducedphysics.org}}
\date{\today}

\begin{document}

\maketitle

%% ---------------------------------------------------------------
%% FRONT-MATTER NOTICES
%% ---------------------------------------------------------------
\noindent\textbf{License.}  Paper text and figures: Creative Commons
Attribution 4.0 (CC BY 4.0).  Source code in the companion
repository: MIT License.

\medskip
\noindent\textbf{Data availability.}  The LaTeX source for this paper
is publicly available at
\href{https://github.com/JackDMenendez/dcl-paper-04-optical-axis-birefringence}%
     {\nolinkurl{github.com/JackDMenendez/dcl-paper-04-optical-axis-birefringence}}.
%% At release time, point to the tag and the Zenodo DOI:
%% The exact commit corresponding to this release is the vX.Y tag,
%% archived at
%% \href{https://doi.org/10.5281/zenodo.XXXXXXXX}%
%%      {\nolinkurl{doi:10.5281/zenodo.XXXXXXXX}}.

\medskip
\noindent\textbf{Keywords.}  A=1 Discrete Causal Lattice; optical-axis
birefringence; kinematic birefringence; gauge-sector birefringence;
$(1,1,-1)$ optical axis; Dirac structure; lattice anisotropy;
multi-channel concordance.

\newpage

\begin{abstract}
    \input{sections/abstract}
\end{abstract}

%% ---------------------------------------------------------------
%% RELEASE NOTES (only at release time)
%% ---------------------------------------------------------------
%% \bigskip
%% \noindent\textbf{Release notes for vX.Y (since vX.Y-1).}
%% \begin{itemize}
%% \item Item 1.
%% \item Item 2.
%% \end{itemize}

%% \emph{Status legend used throughout.}  \texttt{PASS} / \texttt{PART}
%% / \texttt{STUB} / \texttt{FAIL} -- see Table~\ref{tab:audit}.

\newpage
\tableofcontents
\newpage

%% ============================================================
%% PART I: <name>
%% ============================================================
\section{Introduction}
\input{sections/introduction}

\section{Kinematic Dispersion Anisotropy along $(1,1,-1)$}
\input{sections/kinematic_channel}

\section{Gauge-Sector Response and the Birefringence Cancellation along $(1,1,-1)$}
\input{sections/gauge_sector}

\section{Falsifiability: A Multi-Channel Map}
\input{sections/falsifiability}

%% Add more \section{...} \input{sections/...} blocks as needed.
%% Group with comment dividers (e.g., "PART II: DYNAMICS") to keep
%% main.tex skimmable as the paper grows.

\section{Conclusion}
\input{sections/conclusion}

\section*{Acknowledgements}
\input{sections/acknowledgements}

%% ============================================================
%% APPENDICES
%% ============================================================
\appendix

\section{Code and Data}
\input{sections/code_and_data}

\section{Reproducibility Quickstart}
\input{sections/reproducibility}

\section{The Ward-Safe Band-Sum}
\input{sections/band_sum}

%% ============================================================
\bibliographystyle{unsrt}
\bibliography{paper-bib/references}

\end{document}
